---
title: "EDS223-HW4"
author: "Takeen Shamloo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
---

# HW4: Prioritizing Potential Aquaculture

This document analyzes potential aquaculture regions based on sea surface temperature (SST) and depth constraints. We use oyster data as a case study and later generalize the workflow.

## Load Packages

```{r}
suppressMessages(library(here))
suppressMessages(library(terra))
suppressMessages(library(tidyverse))
suppressMessages(library(tmap))
suppressMessages(library(kableExtra))
suppressMessages(library(sf))
```

## Load Data

```{r}
set.seed(123)

# Load SST raster stack
sst_raster_stack <- rast(list.files(
  path = here("data"),
  pattern = "average_annual_sst_.*\\.tif$",
  full.names = TRUE
))

# Load bathymetry raster
bathymetry_raster <- rast(here("data", "depth.tif"))

# Load EEZ shapefile
eez_shape <- vect(here("data", "wc_regions_clean.shp"))
```

## Data Preparation

### Reproject Data

```{r}
# Define UTM Zone 10N CRS
utm_crs <- "+proj=utm +zone=10 +datum=WGS84 +units=m +no_defs"

# Reproject datasets
sst_raster_projected <- project(sst_raster_stack, utm_crs)
bathymetry_projected <- project(bathymetry_raster, utm_crs)
eez_shape_projected <- project(eez_shape, utm_crs)
```

### Calculate Mean SST and Convert to Celsius

```{r}
# Mean SST (2008-2012)
sst_mean <- mean(sst_raster_projected)
sst_mean_celsius <- sst_mean - 273.15  # Convert from Kelvin to Celsius
```

### Prepare Depth Data

```{r}
# Align bathymetry raster with SST extent and resolution
bathymetry_cropped <- crop(bathymetry_projected, ext(sst_mean_celsius))
bathymetry_resampled <- resample(bathymetry_cropped, sst_mean_celsius, method = "near")
```

## Suitability Analysis for Oysters

### Reclassify SST and Depth

```{r}
# Define suitability ranges for oysters
min_sst <- 11  # Minimum SST in Celsius
max_sst <- 30  # Maximum SST in Celsius
min_depth <- 0  # Minimum depth in meters
max_depth <- 70  # Maximum depth in meters

# SST suitability classification
sst_reclass_matrix <- matrix(
  c(-Inf, min_sst, 0, min_sst, max_sst, 1, max_sst, Inf, 0),
  ncol = 3, byrow = TRUE
)
sst_suitability <- classify(sst_mean_celsius, rcl = sst_reclass_matrix)

# Depth suitability classification
depth_reclass_matrix <- matrix(
  c(-Inf, min_depth, 0, min_depth, max_depth, 1, max_depth, Inf, 0),
  ncol = 3, byrow = TRUE
)
depth_suitability <- classify(bathymetry_resampled, rcl = depth_reclass_matrix)
```

### Zonal Analysis

```{r}
# Combine SST and Depth suitability
combined_habitat_suitability <- sst_suitability * depth_suitability

# Rasterize EEZ shapefile for zonal analysis
eez_raster <- rasterize(eez_shape_projected, combined_habitat_suitability, field = "rgn_id")

# Summarize total suitable area by EEZ region
habitat_suitability_by_eez <- zonal(combined_habitat_suitability, eez_raster, fun = "sum", na.rm = TRUE)

# Rename columns for clarity
colnames(habitat_suitability_by_eez) <- c("region_id", "total_suitable_area")
```

### Visualize Suitability

```{r}
# SST Suitability Map
plot(sst_suitability, main = "SST Suitability", col = c("white", "darkgreen"))

# Depth Suitability Map
plot(depth_suitability, main = "Depth Suitability", col = c("white", "blue"))

# Combined Suitability Map
plot(combined_habitat_suitability, main = "Combined Habitat Suitability", col = c("white", "red"))
#plot(eez_shape_projected, add = TRUE, border = "black", lwd = 0.1)

print("Zonal Analysis Results:")
print(habitat_suitability_by_eez)
```

## General Function for Suitability Analysis

```{r}
calculate_habitat_suitability <- function(sst_stack, bathymetry, eez_shapefile, 
                                          sst_min, sst_max, depth_min, depth_max, species_name) {
  # Reproject data to UTM Zone 10N CRS
  utm_crs <- "+proj=utm +zone=10 +datum=WGS84 +units=m +no_defs"
  sst_projected <- project(sst_stack, utm_crs)
  bathymetry_projected <- project(bathymetry, utm_crs)
  eez_projected <- project(eez_shapefile, utm_crs)
  
  # Calculate mean SST and convert to Celsius
  sst_mean <- mean(sst_projected)
  sst_celsius <- sst_mean - 273.15
  
  # Crop and resample bathymetry raster
  bathymetry_cropped <- crop(bathymetry_projected, ext(sst_celsius))
  bathymetry_resampled <- resample(bathymetry_cropped, sst_celsius, method = "near")
  
  # Reclassify SST suitability
  sst_reclass <- matrix(
    c(-Inf, sst_min, 0, sst_min, sst_max, 1, sst_max, Inf, 0),
    ncol = 3, byrow = TRUE
  )
  sst_suitability <- classify(sst_celsius, rcl = sst_reclass)
  
  # Reclassify depth suitability
  depth_reclass <- matrix(
    c(-Inf, depth_min, 0, depth_min, depth_max, 1, depth_max, Inf, 0),
    ncol = 3, byrow = TRUE
  )
  depth_suitability <- classify(bathymetry_resampled, rcl = depth_reclass)
  
  # Combine SST and depth suitability
  combined_suitability <- sst_suitability * depth_suitability
  
  # Perform zonal analysis
  eez_raster <- rasterize(eez_projected, combined_suitability, field = "rgn_id")
  zonal_results <- zonal(combined_suitability, eez_raster, fun = "sum", na.rm = TRUE)
  
  # Rename columns for clarity
  colnames(zonal_results) <- c("region_id", "total_suitable_area")
  
  # Plot SST Suitability
  plot(
    sst_suitability,
    main = paste(species_name, "SST Suitability"),
    col = c("white", "darkgreen"),
    legend.args = list(text = "Suitability", side = 4, line = 2.5)
  )
  
  # Plot Depth Suitability
  plot(
    depth_suitability,
    main = paste(species_name, "Depth Suitability"),
    col = c("white", "blue"),
    legend.args = list(text = "Suitability", side = 4, line = 2.5)
  )
  
  # Plot Combined Suitability
  plot(
    combined_suitability,
    main = paste(species_name, "Combined Suitability"),
    col = c("white", "red"),
    legend.args = list(text = "Suitability", side = 4, line = 2.5)
  )
  #plot(eez_projected, add = TRUE, border = "black", lwd = 0.1)
  
  # Print and return results
  print(paste(species_name, "Habitat Suitability Results:"))
  print(zonal_results)
  return(zonal_results)
}
```

```{r}
# Parameters for Squid
species_name <- "Squid"
suitable_sst_min <- 8  # Minimum SST for squid
suitable_sst_max <- 13 # Maximum SST for squid
suitable_depth_min <- 20  # Minimum depth for squid
suitable_depth_max <- 180 # Maximum depth for squid

# Run Suitability Function
squid_suitability <- calculate_habitat_suitability(
  sst_stack = sst_raster_stack,
  bathymetry = bathymetry_raster,
  eez_shapefile = eez_shape,
  sst_min = suitable_sst_min,
  sst_max = suitable_sst_max,
  depth_min = suitable_depth_min,
  depth_max = suitable_depth_max,
  species_name = species_name
)
```

## Reflection

This assignment was all about figuring out where oysters could live based on 
their habitat needs. Using sea surface temperature (SST) and depth data, 
I calculated where the conditions were suitable. The results showed that about 
93.9% of the area had the right SST (11–30°C), which wasn't surprising since 
most coastal waters meet this range. However, only 2.1% of the area had the 
right depth (0–70m), making the combined suitability really sparse at just 
0.057% of the total area.

Even though the combined suitability was low, this makes sense for oysters, 
since they live in shallow coastal areas. It worked well to highlight 
the regions along the coast that meet both SST and depth conditions. These 
areas are the most promising for oyster aquaculture.

We also made a generalized function to make the same analysis for other species. 
For example, when I used the ranges for squid (SST: 8–13°C, Depth: 20–180m), 
the results showed different suitable areas, which makes sense because squid 
have different habitat needs.